---
title: "Assignment: Lexicon-Based Sentiment Analysis with Bluesky Data"
author: "Aakriti Poudel"
date: "April 24, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

## Overview

In this lab we'll collect posts from Bluesky Social using the `{atrrr}` package and apply lexicon-based sentiment analysis tools — the same tools we covered in lecture (VADER via `{vader}`, `{sentimentr}`, and `{tidytext}` lexicons). By the end you'll have a working pipeline from API call to sentiment scores, and you'll have a concrete sense of where these tools succeed and where they fall short. We'll walk through some of this in class with sample data, then you'll be responsible for completing it with your own data set for the assignment.

---

## Setup

### 1. Create a Bluesky account and app password (you'll do this your assignment)

I created a Bluesky account at bsky.app. To get API access, I went to **Settings - Privacy and Security - App Passwords**, and created a new app password named "eds231-text-sentiment-analysis", and copied the password it gave me.

I saved my credentials in `.Renviron` by running this in the console.

```{r renviron, eval=FALSE}
# Run once in console (not in the Rmd)
usethis::edit_r_environ()
```

### 2.Packages

```{r install, eval=FALSE}
# Install required packages
# install.packages(c("atrrr", "tidyverse", "tidytext", "sentimentr", "remotes", "devtools"))

# tidyvader isn't on CRAN:
# remotes::install_github("chris31415926535/tidyvader")
```

```{r load}
# Load all necessary package
library(atrrr)
library(tidyverse)
library(tidytext)
library(sentimentr)
library(remotes)
library(tidyvader)
library(textdata)

```

## Part 1: Authenticate and Pull Posts

To connect to Bluesky, I authenticated using the credentials I stored in  `.Renviron`. I only need to run this once per session.

```{r auth, eval=FALSE}
auth(
  user = Sys.getenv("BSKY_USER"),
  pass = Sys.getenv("BSKY_PASS")
)
```

We search for posts on an environmental keyword. We'll use `"climate change"` as the default.

```{r search, eval=FALSE}
# API call — only run this manually in the console
raw_posts <- search_post("climate change", limit = 1000) 
```

```{r load-data}
# Load data from CSV instead
raw_posts <- read.csv("https://raw.githubusercontent.com/MaRo406/EDS-231-text-sentiment/refs/heads/main/data/bluesky_climate_raw.csv")
```

Take a look at what came back:

```{r explore}
glimpse(raw_posts)
```

The key columns for our purposes are `text` (post content) and `uri` (unique identifier), and engagement metrics like `like_count` and `reply_count`.

---

## Part 2: Clean and Prepare Text

Pull out the columns we need and do basic cleaning.

```{r clean}
posts <- raw_posts %>%
  select(uri, indexed_at, text, like_count, reply_count, repost_count) %>%
  mutate(
    indexed_at = as.POSIXct(indexed_at), #who can tell me how to 
    post_date  = as.Date(indexed_at),
    text_clean = str_remove_all(text, "https?\\S+"), #remove URLs
    text_clean = str_remove_all(text_clean, "@\\S+"),#remove mentions
    text_clean = str_replace_all(text_clean,"\u2019", "'"),
    text_clean = str_remove_all(text_clean, "'s\\b"),
    text_clean = str_squish(text_clean)) %>%
  filter(nchar(text_clean) > 10)
```

How many posts do we have?

```{r count}
nrow(posts)
```

---

## Part 3: Word Frequency

Before running any sentiment analysis, look at what words are actually in the data and plot the most frequent.

```{r word-freq}
# Pass 1: Standard stop words
data("stop_words")

tokens_clean <- posts %>%
  unnest_tokens(word, text_clean) %>%
  anti_join(stop_words, by = "word")

# Pass 2: Custom stop words and additional cleaning
custom_sw <- tibble(word = c(
  # search terms
  "climate", "change"
  # other common but uninformative text?
))

tokens_clean2 <- tokens_clean %>%
  anti_join(custom_sw, by = "word")

# Word frequency plot
word_counts <-tokens_clean2 %>%
  count(word, sort =TRUE)

word_counts %>%
  slice_max(n, n=20) %>%
  ggplot(aes(x=reorder(word, n), y=n)) +
           geom_col() +
           coord_flip() +
            labs(title = "Top 20 words in Blue Sky climate change posts",
                 x =NULL, y = "Count")
```


## Part 4: Sentiment Analysis

> **Note:** If you need to re-run any chunk in this section, run `rm(posts)` in the console first and re-run from Part 2. This avoids duplicate column errors.

### 4a. Bing lexicon (word-level)

Use `get_sentiments("bing")` to load the lexicon, then tokenize with `unnest_tokens()`, join to the lexicon with `inner_join()`, and compute a net score (positive word count minus negative word count) per post. Join the scores back onto `posts` with `left_join()`.

```{r bing}
bing <- get_sentiments("bing")

# Score posts with Bing
bing_scores <- posts %>% 
  unnest_tokens(word, text_clean) %>%
  inner_join(bing, by = "word") %>%
  count(uri, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate( bing_score = positive - negative)

# Join scores to posts  
posts <- posts %>% 
  left_join(bing_scores, "uri") %>%
  mutate(bing_score = replace_na(bing_score, 0 ))
```

### 4b. AFINN lexicon (numeric scores)

Load with `get_sentiments("afinn")`. Unlike Bing, each matched word carries an integer value (−5 to +5), so summarize by summing those values within each post using `group_by()` and `summarise()`.

```{r afinn}
# You will be prompted to download the AFINN lexicon the first time —
# run this line in the console and select Yes, then it will be cached locally
afinn <- get_sentiments("afinn")

# Score text (break each post into words, match to afinn, sum the scores per post)
afinn_scores <- posts %>%
  unnest_tokens(word, text_clean) %>%
  inner_join(afinn, by = "word") %>%
  group_by(uri) %>%
  summarise(afinn_score = sum(value))

# Join scores to posts, fill missing with 0 (posts with no matched words)
posts <- posts %>%
  left_join(afinn_scores, by = "uri") %>%
  mutate(afinn_score = replace_na(afinn_score, 0))
```

### 4c. NRC lexicon (emotion categories)

The NRC lexicon tags words with 8 emotions (anger, fear, anticipation, trust, surprise, sadness, joy, disgust) plus positive/negative. Now  count emotion words per post and visualize which emotions dominate the corpus.

```{r nrc}
nrc <- get_sentiments("nrc")

# Break posts into words and match to NRC emotion categories
nrc_scores <- posts %>%
  unnest_tokens(word, text_clean) %>%
  inner_join(nrc, by = "word", relationship = "many-to-many") %>%
  filter(!sentiment %in% c("positive", "negative"))

# Count how many emotion words each post has (one row per post per emotion)
nrc_post_scores <- nrc_scores %>%
  count(uri, sentiment)

# Join back to posts (wide format, one column per emotion)
posts <- posts %>%
  left_join(nrc_post_scores %>%
              pivot_wider(names_from = sentiment, values_from = n, values_fill = 0,
                          names_prefix = "nrc_"), by = "uri")
```

Which emotions dominate the corpus? Plot the count of total scored words for each of the 8 emotions.

```{r nrc-plot}
# Count total words in each emotion category across all posts
nrc_scores %>%
  count(sentiment) %>%
  ggplot(aes(x = reorder(sentiment, n), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Total emotion words in Bluesky climate change posts",
       x = NULL,
       y = "Word count") +
  theme_classic()
```

What words are driving each emotion? Create plots of the top 10 words for each emotion (hint: ggplot and `facet_wrap(~sentiment)`).

```{r nrc-top-words}
# Find the top 10 words driving each emotion
posts %>%
  unnest_tokens(word, text_clean) %>%
  inner_join(nrc, by = "word", relationship = "many-to-many") %>%
  filter(!sentiment %in% c("positive", "negative")) %>%
  count(sentiment, word, sort = TRUE) %>%
  group_by(sentiment) %>%
  slice_max(n, n = 10) %>%
  ungroup() %>%
  ggplot(aes(x = reorder_within(word, n, sentiment), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  scale_x_reordered() +
  coord_flip() +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(
    title = "Top 10 words for each emotion in climate change posts",
    x = NULL,
    y = "Counts") +
  theme_minimal()
```

### 4d. sentimentr (sentence-aware)

Unlike the word-level lexicons above, `{sentimentr}` scores at the sentence level and handles valence shifters like negation — "not alarming" won't score the same as "alarming." Pass the full `text_clean` vector to `sentiment_by()`. Rename the resulting `ave_sentiment` to `sr_scores` and join back onto `posts` by row index. I'll give you this code as we haven't fully covered this yet.

```{r sentimentr}
# Split text into sentences first, then score each post
sr_scores <- sentiment_by(get_sentences(posts$text_clean)) %>%
  select(element_id, ave_sentiment) %>%
  rename(sr_score = ave_sentiment) %>%
  mutate(element_id = as.integer(element_id))

# Join scores back to posts by row index
posts <- posts %>%
  mutate(element_id = row_number()) %>%
  left_join(sr_scores, by = "element_id")
```

### 4e. VADER

VADER scores the full post text (no tokenization needed) and returns four values: `pos`, `neu`, `neg`, and `compound`. Use `tidyvader::vader()` on the `text_clean` column — it mutates the result directly onto the dataframe. Keep only the `compound` score and drop the component columns.

Let's run VADER on a single post manually before applying it to the whole dataset:

```{r vader-demo}
# Look at a few posts
posts %>% select(text_clean) %>% slice(1:10)

# Run VADER on one post, swap in whichever row looks interesting to you
posts$text_clean[3]
tidyvader::vader_chr(posts$text_clean[3])
```

What did VADER catch? What did it miss? How does the compound score relate to the word-level scores?

Now score all the posts with VADER.  Rename it's default scoring column to `vader_score`.
---
```{r vader}
# Run VADER on each post and keep only the compound score
posts <- posts %>%
  select(-any_of(c("vader_score", "pos", "neu", "neg"))) %>%  # drop if already exists from a previous run
  tidyvader::vader(text_clean) %>%
  rename(vader_score = compound) %>%
  select(-pos, -neu, -neg)
```

---

## Part 5: Compare the Methods

### Correlations 
Let's see to what extent the 4 numerical methods (bing, afinn, sr, and vader) are in agreement. Select the 4 scores and use `cor()` using "complete.obs".

```{r correlations}
# Check how much the four methods agree with each other
posts %>%
  select(any_of(c("bing_score", "afinn_score", "sr_score", "vader_score"))) %>%
  cor(use = "complete.obs") %>%
  round(2)
```

### Score distributions

Use `pivot_longer()` to reshape the four score columns into a single `score` column with a `method` identifier, then plot with `geom_histogram()` and `facet_wrap()`.

```{r distributions}
# Reshape scores from wide to long so each method gets its own panel
posts %>%
  select(uri, bing_score, afinn_score, sr_score, vader_score) %>%
  pivot_longer(
    cols = c(bing_score, afinn_score, sr_score, vader_score),
    names_to  = "method",
    values_to = "score") %>%
  ggplot(aes(x = score, fill = method)) +
  geom_histogram(bins = 30, show.legend = FALSE) +
  facet_wrap(~method, scales = "free_x") +  # free x so each method uses its own scale
  labs(title = "Score distributions across sentiment methods",
       x = "Sentiment score",
       y = "Number of posts")
```

**Question 1:** Do the four methods broadly agree or disagree? Where do you see the biggest differences in the distributions? Which method produces the most zeros, and why?

**Answer:** The four methods mostly agree. They all lean slightly negative or neutral, but they differ in shape.

`AFINN` spreads out the most, with scores ranging from about -10 to +10. This is because it adds up word scores across the whole post, so longer posts with more emotional words get bigger totals. `Bing` is similar but more compressed. It only counts positive vs. negative words without weighting them.

`sentimentr` and `VADER` both stay close to zero and look more like bell curves. They score the full sentence rather than just matched words, so neutral text pulls the score toward the middle instead of leaving it at zero.

`Bing` produces the most zeros. When a post has no words that match the Bing list at all, which happens often with political or scientific language, the score comes out as zero by default. The tall bar at zero in the `Bing` panel shows this clearly.

---

## Part 6: Spot-Check for Misclassification

Find posts where the methods disagree most — these are the best candidates for understanding failure modes. Use `mutate()` to create binary flags from the continuous scores, then `filter()` for posts where the two methods point in opposite directions.

```{r disagreements}
# Find posts where VADER and Bing point in opposite directions
posts %>%
  mutate(vader_pos = vader_score > 0,  # TRUE if VADER says positive
         bing_pos  = bing_score  > 0   # TRUE if Bing says positive
         ) %>%
  filter(vader_pos != bing_pos) %>%  # keep only posts where they disagree
  select(text_clean, bing_score, vader_score) %>%
  slice_head(n = 10)
```

To rank posts by total disagreement, compute the absolute difference between the two scores with `abs()`, then `arrange(desc(...))` to surface the highest-conflict posts.

```{r high-conflict}
# Rank posts by how much VADER and Bing disagree
posts %>%
  mutate(
    # Normalize bing to [-1, 1] so it's on the same scale as vader, then take the difference
    conflict = abs(vader_score - (bing_score / max(abs(bing_score), na.rm = TRUE)))) %>%
  arrange(desc(conflict)) %>%        # put the biggest disagreements at the top
  select(text_clean, bing_score, afinn_score, sr_score, vader_score) %>%
  slice_head(n = 10)                 # show the 10 most conflicted posts
```

**Question 2:** Read through 5 of the high-conflict posts. For each one, write a sentence explaining which method you think is closer to correct and why.

**Answer:**

  1. *"Do you think the frozen conflict could stall climate change? Perhaps it could act as a third pole. We've lost the Arctic, but hell is freezing over on the Middle East. Hurrah, we are saved."*
  - `VADER` is closer to wrong here, and so is Bing. This is sarcasm. "Hurrah" and "saved" are positive words, but the post is clearly not positive. Neither method catches this, but Bing scores it less positive than `VADER` because it doesn't put as much weight on words like "Hurrah" and "saved."
  
  2. *"Climate change isn't real (sarcasm making sure that obvious lol)"*
  - Neither method does well, but `VADER` is slightly closer because it gives a lower positive score than `Bing`. The word "real" is not in Bing's list as a positive or negative word, so `Bing` scores it close to zero. The post is really just someone making a joke, so a near zero score is actually closer to right.
  
  3. *"If you care about human health or antimicrobial resistance, risk of pandemic disease, plastic pollution, deforestation, loss of biodiversity, climate change..."*
  - This post lists a bunch of serious problems, so a negative score makes sense. Bing picks up words like "risk" and "pollution" and scores it negative, which feels right. `VADER` scores it closer to neutral because there are no exclamation marks or strong emotional words. It just reads like someone rattling off a list.
  
  4. *"yeah except the renewables industry is pushing the idea that we NEED this huge data center demand..."*
  - This post sounds frustrated but doesn't use many clearly positive or negative words, so `Bing` scores it near zero. `VADER` is closer here because it picks up on the all caps "NEED" as a sign of strong feeling, which is exactly what `VADER` was designed to catch in social media text.
  
  5. *"Honestly, every lobby is getting this world destroyed! Their lust for money and power is hurting humanity as a whole!"*
  - This post is clearly negative and angry. Words like "destroyed" and "hurting" are strong. The two exclamation marks add to that. `VADER` is closest here because it was built for this kind of writing short, emotional and punchy social media text.

**Question 3:** Based on everything you've run, which method would you choose if you were delivering a report on public sentiment around your topic, and why? What additional steps would you take to validate your choice?

**Answer:** I would use `VADER`. Climate change posts on Bluesky are short and emotional, and `VADER` was built for that kind of text. It also scored every post, not just ones with matched words like Bing does. The correlation table backs this up. `VADER` and `AFINN` agreed the most (0.80), which suggests `VADER` is tracking real sentiment.

To validate, I would hand label around 100 posts myself and see how often `VADER` agrees. I would also read through the posts where `VADER` and `sentimentr` disagree most to see which one got it right.

---

## Bonus: Engagement vs. Sentiment (completion optional)

Do more emotionally charged posts get more engagement? Use `abs()` on the VADER compound score to capture emotional intensity regardless of direction, apply `log1p()` to `like_count` to handle the skewed distribution, and add `geom_smooth(method = "lm")` to visualize the trend.

```{r engagement}
# Do more emotional posts get more likes?
posts %>%
  filter(like_count > 0) %>% # remove posts with no likes
  mutate(
    sentiment_intensity = abs(vader_score), # intensity regardless of positive or negative
    log_likes = log1p(like_count) # log scale to handle posts with very high likes
  ) %>%
  ggplot(aes(x = sentiment_intensity, y = log_likes)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", formula = y ~ x) +
  labs(title = "Does stronger sentiment lead to more likes?",
       x = "Sentiment intensity (absolute VADER score)",
       y = "Log likes")
```

**Bonus:** Is there a relationship between sentiment intensity and engagement? What does that suggest about what gets amplified on Bluesky?

**Answer:**

The trend line is nearly flat, so there is no clear relationship between sentiment intensity and likes. Emotional posts do not consistently get more likes than neutral ones on Bluesky.

This suggests that on Bluesky, what gets attention is not just how emotional a post is. Other things like who posted it, how many followers they have, or whether it was reposted likely matter more than sentiment alone.

---




