---
title: 'Week 6: LDA Topic Modeling with Nexis Uni Data'
author: "Aakriti Poudel"
date: May 08, 2026
format:
  html:
    toc: true
    toc-float: true
    theme: flatly
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "/Users/aakriti/Documents/MEDS/EDS 231/text-sentiment-analysis")
```

## About the Data

The data for this analysis comes from [Nexis Uni](https://www.library.ucsb.edu/), a news database
available through the UCSB Library. To collect the articles, I searched the keyword **"wild fire"** in
the News category and selected **100 articles** for download in Word (.DOCX) format.

The search was not restricted to a specific country, date range, or news source, so the corpus includes
articles from a range of international outlets. This matters for interpreting the topics -- because the
corpus is global rather than focused on a single region, the model surfaces geographic subthemes
such as Scottish fire response or Nigerian fire incidents as separate topics, rather than grouping all
articles under one broad wildfire theme.

The final corpus contains **100 articles** (as shown in the `glimpse()` output below).

------------------------------------------------------------------------

## Background: Latent Dirichlet Allocation

**Latent Dirichlet Allocation (LDA)** is a method for discovering hidden themes (called *topics*) in a collection of text documents. Each document is treated as a mix of topics, and each topic is a mix of words. The model works backwards from the words it can see to figure out what topics were likely involved. Understanding the key variables helps interpret the results:

| Symbol | Name | What it means |
|-----------------------------|----------------------|----------------------|
| **α** (alpha) | Hyperparameter | Controls how many topics a single document tends to cover. A small value means each document focuses on just one or two topics. Set by the researcher. |
| **β** (beta) | Hyperparameter | Controls how many words each topic tends to use. A small value means each topic has a tight, focused vocabulary. Set by the researcher. |
| **θ** (theta) | Document-topic mix | The estimated proportion of each topic in a given document -- i.e., how much of document X is about topic Y. Has to be estimated from the data. |
| **φ** (phi) | Topic-word mix | The estimated probability of each word belonging to a given topic -- i.e., how strongly word W is associated with topic T. Called `beta` in tidytext. Has to be estimated from the data. |
| **z** | Topic assignment | The hidden topic label attached to each individual word. Connects the document-topic and topic-word distributions. Has to be estimated from the data. |
| **w** | Observed word | The only thing you actually see. Everything else (z, θ, φ) is estimated from patterns in w. |

### How the Model Imagines Text Was Written

LDA assumes that documents were written as follows:

1.  For each topic *t*, draw a word distribution **φ_t** \~ Dirichlet(β)
2.  For each document *m*, draw a topic distribution **θ_m** \~
    Dirichlet(α)
3.  For each word position in that document:
    -   Draw a topic assignment **z** \~ Multinomial(θ_m)
    -   Draw the word **w** \~ Multinomial(φ_z)

In practice, we only observe the words (**w**). The model runs this process in reverse to estimate the most likely **z**, **θ**, and **φ**.

### The Dirichlet Distribution

The Dirichlet distribution produces probability vectors that sum to 1, ideal for representing mixtures of topics or words. The concentration parameter (α or β) controls how spread out the mixture is:

-   **Small value (\< 1)**: most probability sits on one or two categories, realistic for most text.
-   **Large value (\> 1)**: probability is spread evenly across all categories.
-   **Value = 1**: all mixtures are equally likely.

------------------------------------------------------------------------

## Setup

```{r packages}
library(LexisNexisTools)
library(quanteda)
library(tm)
library(topicmodels)
library(topicdoc)
library(tidyverse)
library(tidytext)
library(reshape2)
library(tictoc)
library(LDAvis)
library(tsne)
library(servr)
library(slam)
```

## Load the Data

We use `{LexisNexisTools}` to parse the Nexis Uni `.DOCX` export. This package understands the specific formatting Nexis Uni uses and splits the file into separate articles with metadata like date, headline, and source.

```{r data}
# Set path to the Nexis Uni data file
pre_files <- "assignments/data/nexis_uni_wild-fire_data.DOCX"

# Parse the Nexis Uni export into structured slots
pre_dat <- lnt_read(pre_files)

# Extract metadata, article text, and paragraph-level data
pre_meta_df       <- pre_dat@meta
pre_articles_df   <- pre_dat@articles
pre_paragraphs_df <- pre_dat@paragraphs

# Build a clean data frame with one row per article
tbl <- tibble(
  date     = pre_meta_df$Date,
  headline = pre_meta_df$Headline,
  id       = pre_meta_df$ID,
  text     = pre_articles_df$Article,
  source   = pre_meta_df$Newspaper
)

glimpse(tbl)
```

## Build the Corpus

We convert the data frame into a quanteda corpus object. Each row in `tbl` becomes one document in the corpus, which LDA will treat as a separate unit when estimating topic mixtures.

```{r corpus}
corpus <- corpus(x = tbl, text_field = "text")
```

## Tokenize and Clean

Before modeling, we need to break the text into individual words (tokens) and remove words that carry no meaningful information about topic differences. We remove standard English stop words, common journalistic filler phrases, and generic wildfire terms. Words like "fire" and "wildfire" appear in nearly every article and would show up at the top of every topic without helping us understand what makes the topics different from each other. We also remove a stray backtick character (`` ` ``) that can slip through as a tokenization artifact.

```{r tokens}
# Combine standard stopwords with domain-specific ones
add_stops <- c(quanteda::stopwords("en", source = "smart"),
               "said", "says", "told", "according",
               "fire", "fires", "wildfire", "wildfires",
               "`")

# Tokenize, remove punctuation and numbers, then lowercase
toks <- tokens(corpus, remove_punct = TRUE, remove_numbers = TRUE) %>%
  tokens_tolower()

# Remove all stopwords
toks1 <- tokens_select(toks, pattern = add_stops, selection = "remove")
```

## Build the Document-Feature Matrix

LDA works on a document-feature matrix (DFM), which counts how many times each word appears in each document. We remove very rare words (appearing in fewer than 2 documents) since they add noise without helping identify shared topics. We also remove any documents that end up empty after trimming.

```{r dfm}
dfm1 <- dfm(toks1)

# Remove words that appear in fewer than 2 documents
dfm2 <- dfm_trim(dfm1, min_docfreq = 2)

# Remove empty documents -- LDA cannot handle rows with zero tokens
sel_idx <- slam::row_sums(dfm2) > 0
dfm     <- dfm2[sel_idx, ]
```

------------------------------------------------------------------------

## Step 1: Initial Model (Theory-Driven k)

Before using data to choose the number of topics, we start with a theory driven guess. For a wildfire news corpus, we might expect articles to cluster around a few broad themes: emergency response, climate and weather conditions, land management, and property damage. This gives us a starting point of k = 4.

```{r LDA_k4}
k <- 4

topicModel_k4 <- LDA(dfm,
                     k,
                     method  = "Gibbs",
                     control = list(iter    = 100,
                                    verbose = 25,
                                    seed    = 321))
```

### Inspect the Posterior Distributions

`posterior()` extracts the two key output matrices:

-   **theta (θ)**: document × topic matrix, how much of each document belongs to each topic
-   **beta/phi (φ)**: topic × term matrix, how strongly each word is associated with each topic

```{r inspect_k4}
result <- posterior(topicModel_k4)

beta  <- result$terms
theta <- result$topics

cat("phi dimensions  (topics x vocab):", dim(beta),  "\n")
cat("theta dimensions (docs x topics):", dim(theta), "\n")

# Top 10 terms per topic
terms(topicModel_k4, 10)
```

Looking at the top terms, we can see some early patterns. Topic 1 captures Scottish fire response terms like "rescue" and "scottish", and Topic 4 shows general and unexpected words like "smoke", "cooking", and "pizza". However, topics are quite mixed overall, for example "cooking" and "pizza" appear in Topic 4 alongside general news terms, while "qep" appears in Topic 3 alongside industrial terms like "liquid" and "chemical", which are unrelated to each other. This suggests k = 4 is too few to cleanly separate the themes in this 100-article corpus.

------------------------------------------------------------------------

## Step 2: Selecting k with Coherence Scores

Instead of relying only on theory, we can let the data help us choose k. We fit models for k = 2 through 15 and measure coherence for each. Coherence measures whether the top words in each topic actually appear together in the same articles, a sign that the topic is meaningful and not just a mix of unrelated words.

**Higher coherence (closer to 0) = more meaningful topics.**

```{r find_k}
tic()

# Fit one model for each value of k
k_values <- seq(2, 15, by = 1)

coherence_results <- map_dfr(k_values, function(k) {
  m <- LDA(dfm, k = k, method = "Gibbs",
           control = list(iter    = 300,
                          seed    = 321,
                          verbose = 0))

  # Calculate mean coherence score for this k
  coh <- mean(topic_coherence(m, dfm))
  tibble(k = k, coherence = coh)
})

toc()

# Plot coherence vs k
ggplot(coherence_results, aes(x = k, y = coherence)) +
  geom_line(color = "navy") +
  geom_point(color = "navy", size = 2) +
  geom_vline(xintercept = coherence_results$k[which.max(coherence_results$coherence)],
             linetype = "dashed", color = "red") +
  labs(title    = "Mean topic coherence by number of topics (k)",
       subtitle = "Dashed line = k with highest coherence",
       x        = "Number of topics (k)",
       y        = "Mean coherence score") +
  theme_minimal()

# Full table of coherence scores
coherence_results
```

> **Interpreting the plot:** Look for the peak coherence value, or an "elbow" where adding more topics
> stops improving coherence meaningfully. Also inspect the top terms at two or three candidate k values,
> statistical criteria and human interpretability do not always agree.

The coherence plot peaks at **k = 9** (coherence = -59.0). The curve drops sharply from k = 2 to k = 3, then generally improves from k = 3 through k = 9, after which it becomes more variable. However, k = 9 is still more topics than needed for 100 articles and would likely create several near-identical groups.

We proceed with **k = 6** as a practical balance. While k = 7 (-65.5), k = 8 (-62.4), and the statistical peak k = 9 (-59.0) all have higher coherence scores than k = 6 (-66.1), at those values topics begin splitting closely related content into subtopics that are difficult to tell apart. At k = 6, each topic has a clean, distinct set of terms that points to a recognizable real-world theme.

------------------------------------------------------------------------

## Step 3: Fit the Selected Model

We now fit the final model with k = 6, using more iterations (1000) for better convergence than the exploratory model.

```{r LDA_final}
k <- 6

topicModel_k_select <- LDA(dfm, k,
                           method  = "Gibbs",
                           control = list(iter    = 1000,
                                          verbose = 25,
                                          seed    = 321))

# Extract posterior distributions
tmResult <- posterior(topicModel_k_select)
terms(topicModel_k_select, 10)

theta <- tmResult$topics
beta  <- tmResult$terms
vocab <- colnames(beta)
```

------------------------------------------------------------------------

## Step 4: Visualize Top Terms per Topic

The β (φ) matrix gives us the probability of each word belonging to each topic. Plotting the highest-probability terms for each topic is the main tool for understanding what each topic is actually about.

```{r top_terms_plot}
topics <- tidy(topicModel_k_select, matrix = "beta")

top_terms <- topics |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  arrange(topic, -beta)

top_terms |>
  mutate(term = reorder_within(term, beta, topic, sep = "")) |>
  ggplot(aes(term, beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free_y") +
  scale_x_reordered() +
  coord_flip() +
  labs(title    = "Top 10 terms per topic",
       subtitle = "β = P(term | topic)",
       x        = NULL,
       y        = "β") +
  theme_minimal()
```

------------------------------------------------------------------------

## Step 5: Name the Topics

We assign a label to each topic based on its top words. This step requires reading the terms and thinking about what real-world theme they point to.

```{r topic_names}
topic_words <- terms(topicModel_k_select, 5)

# Assign a descriptive label to each topic
# Order matches the topic numbering from terms(topicModel_k_select, 10):
# Topic 1: rescue, scottish, service, government, training
# Topic 2: conditions, wild, climate, cooking, years
# Topic 3: forest, management, areas, local, community
# Topic 4: service, weather, wild, dry, april (firefighters, crews, heat)
# Topic 5: people, qep, news, home, emergency
# Topic 6: wild, property, system, liquid, chemical
topic_names <- c(
  "Scottish Fire Response",
  "Climate & Weather Conditions",
  "Vegetation & Land Management",
  "Seasonal Fire & Firefighter Response",
  "Disaster Recovery & Community Impact",
  "Industrial Fire Risk & Technology"
)

topic_names
```

------------------------------------------------------------------------

## Step 6: Document-Topic Distributions (θ)

The θ matrix shows what proportion of each document belongs to each topic. A document focused entirely on one theme will have a high θ for that topic and near-zero values for the others. Mixed documents spread their weight across several topics.

```{r topic_dists}
# Select the first 5 documents to examine
example_ids <- c(1:5)
n <- length(example_ids)

example_props <- theta[example_ids, ]
colnames(example_props) <- topic_names

viz_df <- melt(cbind(data.frame(example_props),
                     document = factor(1:n)),
               variable.name = "topic",
               id.vars = "document")

ggplot(data = viz_df, aes(topic, value, fill = document)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip() +
  facet_wrap(~ document, ncol = n) +
  labs(title = "Topic distributions across sample documents",
       subtitle = "θ = P(topic | document)",
       x = NULL,
       y = "Topic proportion")
```

------------------------------------------------------------------------

## Step 7: Interactive Visualization with LDAvis

LDAvis provides an interactive view of two complementary aspects of the model:

  - **Intertopic distance map**: circles represent topics;
      size = proportion of corpus tokens assigned to that topic;
      distance = degree of topic overlap in vocabulary
      
- **Term relevance bar chart**: adjustable by the λ (lambda) parameter

### The λ (lambda) Relevance Parameter

λ controls how words are ranked within a topic:

$$\text{Relevance}(w, t) = \lambda \times P(w|t) + (1 - \lambda) \times \frac{P(w|t)}{P(w)}$$

-   **λ near 1**: ranks words by raw frequency within the topic, highlights common words
-   **λ near 0**: ranks words by how *exclusive* they are to that topic relative to the whole corpus, highlights distinctive words
-   **λ = 0.6**: the recommended starting point (Sievert & Shirley, 2014) that balances both, **use this, not the default λ = 1.0**

```{r LDAvis}
svd_tsne <- function(x) tsne(svd(x)$u)

json <- createJSON(
  phi            = tmResult$terms,
  theta          = tmResult$topics,
  doc.length     = rowSums(dfm),
  vocab          = colnames(dfm),
  term.frequency = colSums(dfm),
  mds.method     = svd_tsne,
  plot.opts      = list(xlab = "", ylab = "")
)

if (interactive()) serVis(json)
```

> **Reading the LDAvis plot:** The intertopic distance map shows 6
> circles of varying sizes. Topic 2 (Climate and Weather Conditions) and
> Topic 4 (Seasonal Fire and Firefighter Response) are among the largest
> circles, meaning they account for the biggest shares of the corpus.
> Topic 3 (Vegetation and Land Management) sits far to the left, well
> separated from the rest, which means it uses very different vocabulary
> from the other topics. Topic 6 (Industrial Fire Risk and Technology)
> sits near the center of the map. Topics 4 and 5 sit close together in
> the lower portion of the map, which makes sense since both cover
> community-level response and locally focused fire content. Topic 1
> (Scottish Fire Response) sits near the center-left and is reasonably
> well separated from the other topics. Overall the separation is good,
> though the overlap between Topics 4 and 5 suggests those two could
> potentially be merged.

------------------------------------------------------------------------

## Interpretation

### 1. Build and Clean Your Corpus

The corpus was built from 100 Nexis Uni news articles retrieved
using the keyword search **"wild fire"** in the News category, with no
geographic or date filters applied. The file `nexis_uni_wild-fire_data.DOCX`
was downloaded for this assignment and parsed using `{LexisNexisTools}`,
covering international outlets.

Custom stop words were added in two groups. First, journalistic filler
words ("said", "says", "told", "according") were removed because they
appear across all articles regardless of topic. Second, the words
"fire", "fires", "wildfire", and "wildfires" were removed because they
appear in nearly every article and would otherwise show up at the top of
every topic, making it impossible to see what is actually different
between them. A stray backtick character was also added to the stop word
list to prevent a punctuation artifact from appearing in the top term lists.

### 2. Select k Using Multiple Criteria

**Theory:** A wildfire news corpus covering international outlets is
likely to contain articles on emergency response, land management,
climate and weather conditions, industrial fire technology, community
recovery, and specific geographic events. This suggests somewhere
between 5 and 7 topics would capture the main themes across 100
articles.

**Optimization Metrics:** The coherence plot peaks at k = 9 (coherence
= -59.0). The curve drops at k = 3, then improves generally through k =
9 before becoming more variable. k = 9 is still more topics than needed
for 100 articles and would likely split closely related content into
near-identical groups. While k = 7 (-65.5), k = 8 (-62.4), and the statistical peak k = 9
(-59.0) all score higher than k = 6 (-66.1), at those values topics
begin splitting closely related content into subtopics that are hard to
tell apart.

**Interpretability:** At k = 4, topics are broad and contain clearly
unrelated terms mixed together, for example "cooking" and "pizza"
appear in Topic 4 alongside weather adjacent terms like "smoke", while
"qep" appears in Topic 3 alongside industrial suppression terms like
"liquid" and "chemical". These are unrelated clusters of content that
k = 4 fails to separate. At k = 6, each topic has a cleaner set of
terms that point to a recognizable theme. At k = 8, some topics would
start to split into very similar subtopics.

**Justification:** k = 6 was selected. The coherence score at k = 6 is
-66.1, which is not the highest, but at k = 7, k = 8, and the
statistical peak k = 9, topics start splitting into subtopics that are
difficult to distinguish from each other in practice. The 6 topics
produced are each interpretable and distinct. The LDAvis map shows good
separation overall, with the exception of Topics 4 and 5 which sit close
together, both covering active fire response and recovery content and
are naturally similar themes.

### 3. Final Model and Top Terms Plot

See Step 4 above, top 10 terms per topic plotted with β weights.

### 4. Document-Topic Distributions

See Step 6 above, θ distributions across 5 sample documents from the
100-article corpus. Document 4 is dominated by the Industrial Fire Risk
and Technology topic, meaning that article focuses almost entirely on
fire suppression systems and technical fire-prevention content. Documents
2, 3, and 5 are each dominated by the Seasonal Fire and Firefighter
Response topic, pointing to active-fire-season coverage. Document 1 is
the most spread across topics, meaning that article touches on several
themes within a single piece.

### 5. Interpret Your Topics

**Topic 1: Scottish Fire Response:**
The top terms are "rescue", "scottish", "service", "government",
"training", "report", "sfrs", "firefighters", "states", and "blazes".
This topic covers articles about Scottish fire services, their
operations, training programs, government reports, and responses to
blazes. "sfrs" refers to the Scottish Fire and Rescue Service. The
topic is geographically distinct because Scotland's fire service
operates under its own national structure, which generates a consistent
set of terms across multiple articles.

**Topic 2: Climate & Weather Conditions:**
The top terms are "conditions", "wild", "climate", "time", "cooking",
"years", "solution", "smoke", "people", and "support". This topic groups
articles that place wildfire in the context of climate and weather, dry
conditions, rising temperatures, and seasonal patterns. "cooking" appears
here because several articles about the Wild Fire pizza restaurant in
Sunderland were pulled into the corpus by the broad search term "wild
fire", those articles have no connection to actual wildfires but share
general vocabulary that landed them in this cluster.

**Topic 3: Vegetation & Land Management:**
The top terms are "forest", "management", "areas", "local", "community",
"vegetation", "controlled", "lives", "county", and "risk". This topic
covers articles about managing land to reduce wildfire risk, controlled
burns, clearing vegetation, protecting local communities and wildlife.
It is the most prevention focused topic and deals with what happens
before a fire starts rather than during or after.

**Topic 4: Seasonal Fire & Firefighter Response:**
The top terms are "service", "weather", "wild", "dry", "april",
"firefighters", "country", "crews", "emergency", and "heat". This topic
groups articles about active fire seasons, crews responding to
emergencies during dry spring and summer months. "april" and "dry"
suggest several articles focus on the seasonal window when fire risk
peaks. It covers frontline response from multiple countries.

**Topic 5: Disaster Recovery & Community Impact:**
The top terms are "people", "qep", "news", "home", "emergency",
"content", "california", "state", "recovery", and "disaster". This topic
groups articles about what happens after a wildfire, evacuation,
recovery efforts, and community rebuilding. "qep" refers to QEP, a
company that donated cleaning supplies to hurricane and wildfire cleanup
efforts, which appeared across multiple articles. "california" and
"state" point to US-based recovery coverage.

**Topic 6: Industrial Fire Risk & Technology:**
The top terms are "wild", "property", "system", "liquid", "chemical",
"network", "anti-fire", "area", "spraying", and "homes". This topic
covers industrial fire suppression technology, chemical spraying
systems, anti-fire patents, and property protection networks. It sits
near the center of the LDAvis map but uses highly technical and
product specific vocabulary that sets it apart from the general
wildfire news coverage in the other topics.